Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-05-28 paper/paper.qmd

We performed permutation feature importance to determine which OTUs were the most
important for model performance (@fig-features).

## Clinical value of severity prediction models

@@

@@

Random forest models were used to examine whether OTU data collected on the day
of diagnosis could classify CDI cases as severe according to four different
definitions of severity.
We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021] implemented in
a custom version of the mikropml Snakemake workflow [@sovacool_mikropml_2023]
for all steps of the machine learning analysis.
We have a full dataset which uses all samples available for each severity
definition, and an intersection dataset which consists of only the samples that
have all four definitions labelled.
The intersection dataset is the most fair for comparing model performance across
definitions, while the full dataset allows us to use as much data as possible
for model training and evaluation.
Datasets were preprocessed with the default options in mikropml to remove
features with near-zero variance and scale continuous features from -1 to 1. <!-- No features had missing values and no features were perfectly correlated -->
During preprocessing, `r preproc_ranges$min_removed` to `r preproc_ranges$max_removed`
features were removed due to having near-zero variance, resulting in datasets
ranging from `r preproc_ranges$min_feats` to `r preproc_ranges$max_feats`
depending on the severity definition.
We randomly split the data into an 80% training and 20% test set and repeated
this 100 times, followed by training models with 5-fold cross-validation.
Model performance was calculated on the test set using the area under the
receiver-operator characteristic curve (AUROC) and the area under the balanced
precision-recall curve (AUBPRC).
Permutation feature importance was then performed to determine which OTUs
contributed most to model performance.
We reported OTUs with a significant permutation test in at least 80 of the 100
models.

@@

::: {#fig-performance fig-cap="**Performance of ML models.**
Area under the receiver-operator characteristic curve (AUROC) for the test sets
and cross-validation folds of the training sets,
and the area under the balanced precision-recall curve (AUBPRC) for the test sets.
Left: models were trained on the full dataset, with different numbers of samples
available for each severity definition.
Right: models were trained on the intersection of samples with all labels
available for each definition. 
Note that the intersection dataset for Attributable and Pragmatic severity have
exactly the same labels, thus an identical performance is expected.
Nearly all pairs of definitions have significantly different performances on the
test set (P < 0.05) except for AUROC and AUBPRC of Attributable vs. Pragmatic on
the intersection dataset (as they are identical), AUROC of Attributable vs.
All-cause on the full dataset, and AUROC of Attributable vs. IDSA on the full
dataset.
<!-- TODO add plots of AUROC and AUBPRC curves -->
"}
![](figures/ml-performance.png)
:::

>>>>

We performed permutation feature importance to determine which OTUs contributed
the most to model performance (@fig-features).

## Estimating the benefits of severity prediction models

@@

decision curve analysis.

@@

Random forest models were used to examine whether OTU data collected on the day
of diagnosis could classify CDI cases as severe according to four different
definitions of severity.
We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021] implemented in
a custom version of the mikropml Snakemake workflow [@sovacool_mikropml_2023]
for all steps of the machine learning analysis.
We have a full dataset which uses all samples available for each severity
definition, and an intersection dataset which consists of only the samples that
have all four definitions labelled.
The intersection dataset is the most fair for comparing model performance across
definitions, while the full dataset allows us to use as much data as possible
for model training and evaluation.
Datasets were preprocessed with the default options in mikropml to remove
features with near-zero variance and scale continuous features from -1 to 1. <!-- No features had missing values and no features were perfectly correlated -->
During preprocessing, `r preproc_ranges$min_removed` to `r preproc_ranges$max_removed`
features were removed due to having near-zero variance, resulting in datasets
ranging from `r preproc_ranges$min_feats` to `r preproc_ranges$max_feats`
depending on the severity definition.
We randomly split the data into an 80% training and 20% test set and repeated
this 100 times, followed by training models with 5-fold cross-validation.
Model performance was calculated on the test set using the area under the
receiver-operator characteristic curve (AUROC) and the area under the balanced
precision-recall curve (AUBPRC).
Permutation importance was then performed to determine which OTUs contributed
most to model performance.
We reported OTUs with a significant permutation test in at least 80 of the 100
models.

@@

::: {#fig-performance fig-cap="**Performance of ML models.**
In the left panels, models were trained on the full dataset, with different numbers of samples
available for each severity definition.
In the right panels, models were trained on the intersection of samples with all labels
available for each definition. 
Note that the intersection dataset for Attributable and Pragmatic severity have
exactly the same labels, thus identical values are expected.
**A)** 
Area under the receiver-operator characteristic curve (AUROC) for the test sets
and cross-validation folds of the training sets,
and the area under the balanced precision-recall curve (AUBPRC) for the test sets.
Nearly all pairs of definitions have significantly different performances on the
test set (P < 0.05) except for AUROC and AUBPRC of Attributable vs. Pragmatic on
the intersection dataset (as they are identical), AUROC of Attributable vs.
All-cause on the full dataset, and AUROC of Attributable vs. IDSA on the full
dataset.
**B)** Receiver-operator characteristic curves for the test sets.
Mean specificity is reported at each sensitivity value.
**C)** Balanced precision-recall curves for the test sets.
Mean balanced precision is reported at each recall value.
"}
![](figures/ml-performance.png)
:::
